---
title: "02-02-datatransf"
output: html_document
editor: 
  markdown: 
    wrap: 72
---

## Intro to `tidyverse()` and `dplyr()`

`dplyr()` is a very powerful data management and transformation package
which works upon *tidy data* with the following set of verbs and
functions:

-   Pick observations by their values (`filter()`).
-   Reorder the rows (`arrange()`).
-   Pick variables by their names (`select()`).
-   Create new variables with functions of existing variables
    (`mutate()`).
-   Collapse many values down to a single summary (`summarise()`).

These can all be used in conjunction with `group_by()` which changes the
scope of each function from operating on the entire dataset to operating
on it group-by-group.

## Basic data operations with `dplyr()`

We will work with the full gapminder data.

```{r}
library(tidyverse)
library(gapminder)

life_expt <- as_tibble(gapminder)

head(life_expt)
glimpse(life_expt)
str(life_expt)
skim(life_expt)
```

------------------------------------------------------------------------

## Pick observations based on row values — `filter()`

```{r}
filter(life_expt, year == 2002, continent == "Asia")

filter(life_expt, year %in% c(1997, 2002, 2007), continent == "Asia")
```

```{r}
life_expt |> 
  filter(year == 2002,
         continent %in% c("Asia", "Africa"))
```

### Student exercise: `filter()`

```{r}
#| eval: false
# Filter life_expt to:
# year == 2007
# continent is either Europe or Americas
```

------------------------------------------------------------------------

## Reorder rows — `arrange()`

```{r}
arrange(life_expt, -gdpPercap)
```

```{r}
life_expt |> 
  filter(year == 2007, 
         continent %in% c("Asia", "Africa")) |> 
  arrange(continent, desc(lifeExp))
```

`%in%` is called the membership operator. It is used for filtering and
logical subsetting. It checks if the elements of the left vector exist
anywhere in the right vector. It returns a logical vector (TRUE/FALSE).

### Student exercise: `arrange()`

```{r}
#| eval: false
# Starting from life_expt:
# Filter to year == 2007
# Arrange by lifeExp from highest to lowest
```

------------------------------------------------------------------------

## Pipes (`|>`) as a workflow tool

```{r}
16 |> sqrt()
16 |> sqrt() |> log2()
```

The pipe sends the result of the left-hand side to the **first
argument** of the function on the right.

------------------------------------------------------------------------

## Select columns — `select()` {#select}

```{r}
select(life_expt, country, year, lifeExp)

select(life_expt, country:lifeExp)
```

```{r}
iris_df <- iris

iris_df |> 
  select(starts_with("Sep")) |> 
  names()
```

The `select()` verb is pretty powerful in `dplyr` due to the helper
functions.

###1. Pattern-Based Helpers

These are the most common tools for filtering columns by name.

-   `starts_with("abc")`: Selects columns that begin with "abc".

    -   Example: `select(df, starts_with("year"))`

-   `ends_with("xyz")`: Selects columns that end with "xyz".

    -   Example: `select(df, ends_with("_id"))`

-   `contains("def")`: Selects columns that contain the string "def"
    anywhere.

    -   Example: `select(df, contains("margin"))`

### 2.  Positional & Logic Helpers 
These help you grab columns based on their order or relationship to other selections.

- `everything()`: Matches all variables. Useful for reordering columns
(e.g., select(df, target_var, everything()) moves target_var to the
first column).

- `last_col()`: Selects the very last column. You can add an offset, like
last_col(offset = 1) for the second-to-last.

- `all_of(vector)`: Selects columns based on a character vector. If a name
is missing, it throws an error.

- `any_of(vector)`: Same as all_of, but it doesn't error if some columns
aren't found. This is safer for flexible scripts.

### 3.  Data-Type Helpers (The `where()` function) 

In modern dplyr, if you want to select based on the type of data (e.g., all numeric columns), you use the where() helper.

- `where(is.numeric)`: Selects all double or integer columns.

- `where(is.character)`: Selects all string columns.

- `where(is.factor)`: Selects all categorical factor columns.

Helper, Use Case starts_with(),"Identifying prefixes (e.g., income\_)"
everything(),Reordering columns where(is.numeric),Selecting features for
math/stats any_of(),Selecting from a list of strings safely :,"Selecting
a range (e.g., col_a:col_f)"

### Student exercise: `select()`

```{r}
#| eval: false
# Using life_expt:
# Select only country, continent, and gdpPercap
```

------------------------------------------------------------------------

## Rename variables — `rename()`

```{r}
life_expt |> 
  rename(gdpPerCapita = gdpPercap,
         population = pop) |> 
  summarise(meanpop = mean(population),
            medgdppc = median(gdpPerCapita))
```

------------------------------------------------------------------------

## Add new variables — `mutate()`

```{r}
life_expt |> 
  mutate(nomGdp = pop * gdpPercap)
```

```{r}
life_expt2 <- life_expt |>  
  mutate(
    nomGdp = pop * gdpPercap,
    ln_pop = log(pop)
  ) |> 
  select(-pop, -gdpPercap)

head(life_expt2)
```

### Student exercise: `mutate()`

```{r}
#| eval: false
# Using life_expt:
# Create a new variable gdp_total = pop * gdpPercap
# Keep only country, year, and gdp_total
```

------------------------------------------------------------------------

## Grouping data — `group_by()` and `top_n()`

```{r}
life_expt |> 
  filter(year == 2007) |> 
  group_by(continent) |> 
  top_n(2, lifeExp) |> 
  arrange(continent, -lifeExp)
```

### Student exercise: grouping

```{r}
#| eval: false
# For year 2007:
# Find the top 3 countries by life expectancy in each continent
```

------------------------------------------------------------------------

## Collapse data — `summarise()`

```{r}
life_expt |> 
  group_by(continent) |> 
  summarise(avLE = mean(lifeExp))
```

```{r}
life_expt |>  
  group_by(continent, year) |>  
  summarise(meanLifeExp = mean(lifeExp, na.rm = TRUE),
            medianLifeExp = median(lifeExp, na.rm = TRUE),
            sdLifeExp = sd(lifeExp, na.rm = TRUE))
```

### Student exercise: `summarise()`

```{r}
#| eval: false
# Group life_expt by continent
# Compute mean and median life expectancy
```

------------------------------------------------------------------------

## Another data management application

```{r}
library(mosaicData)

housing_df <- as_tibble(SaratogaHouses)
glimpse(housing_df)
```

```{r}
housing_df2 <- housing_df |>  
  mutate(pricePerSqF = price / livingArea)
```

```{r}
housing_df3 <- housing_df2 |> 
  mutate(
    size = case_when(
      livingArea <= 1100 ~ "small",
      livingArea <= 2500 ~ "medium",
      .default = "large"
    )
  ) |> 
  mutate(size = factor(size, levels = c("small", "medium", "large")))
```

```{r}
housing_df3 |> 
  group_by(size, newConstruction) |>  
  summarise(median_price = median(price),
            avg_livingArea = mean(livingArea, na.rm = TRUE))
```

------------------------------------------------------------------------

## Student exercises: cumulative `dplyr` practice

```{r}
#| eval: false
# Using housing_df:
# Filter to waterfront == "Yes"
# Compute the average price
```

```{r}
#| eval: false
# Using housing_df:
# Group by waterfront
# Report median price and average livingArea
```

```{r}
#| eval: false
# Using life_expt:
# Group by continent and year
# Compute mean life expectancy
```
