---
title: "02-Feature Selection"
author: "Dr. Jeff Jacob"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Feature Selection

### Learning objectives

After completing this section, you should be able to:

* Explain why feature selection helps reduce overfitting
* Distinguish between forward, backward, and stepwise selection
* Use `caret` to implement sequential feature selection methods
* Compare models using RMSE and parsimony
* Evaluate feature-selected models on a validation set

---

## Motivation: Why feature selection?

Including too many predictors can lead to **overfitting**—a model that fits training data well but performs poorly on new data. Feature selection aims to retain predictors that meaningfully improve prediction accuracy while removing redundant or noisy variables.

Model selection is typically guided by:

* Validation-set RMSE
* Information criteria (AIC, BIC)
* Interpretability and parsimony

There are two broad approaches:

1. **Sequential selection** (add/remove predictors)
2. **Regularization** (penalize complexity)

This chapter focuses on **sequential selection**.

---

## Load the required libraries

```{r}
library(tidyverse)
library(broom)
library(forecast)
library(caret)
library(mlbadata)
library(fastDummies)
```

---

## Data preparation

We continue using the **Toyota Corolla** dataset to ensure comparability with the baseline regression model.

```{r}
car.df <- mlbadata::ToyotaCorolla

# use first 1000 rows
car.df <- car.df[1:1000, ]

# convert fuel type to factor
car.df$Fuel_Type <- as_factor(car.df$Fuel_Type) |> 
  fct_relevel("CNG")
selected.var <- c(3, 4, 7, 8, 9, 10, 12, 13, 14, 17, 18)
car.df.sel <- car.df[, selected.var]

glimpse(car.df.sel)
```

---

### Student exercise 1: intuition check

```{r}
#| eval: false
# Which variables do you expect to be dropped
# during feature selection?
# Which do you expect to survive?
```

---

## Creating full-rank dummy variables

We will first partition the data into training and validation set and then apply pre-processing steps to the training set. We will then apply those steps to the test data with inputs from the training set. This prevents data leakage. For this reason, we will use `caret()` library to create dummy variables and then apply that step to training and test set.




## Train–validation split

```{r}
set.seed(1)
trainInd <- createDataPartition(car.df$Price, p = 0.6, list = FALSE)

train.df <- car.df[trainInd, ]
valid.df <- car.df[-trainInd, ]
```

Now, to avoid perfect multicollinearity, we explicitly create **full-rank dummy variables** by removing one category per factor.


```{r}
dum_vars <- dummyVars(Price ~ .,
                        data = car.df.sel,
                        fullRank = TRUE)
 
```

Now, we apply this object on the training and test set.

```{r}
X_train <- predict(dum_vars, newdata = train.df) |> as.data.frame()
X_valid <- predict(dum_vars, newdata = valid.df) |> as.data.frame()

y_train <- train.df$Price
y_valid <- valid.df$Price
```


---

## Sequential feature selection methods

### Overview

* **Forward selection**: start with intercept, add predictors
* **Backward elimination**: start with full model, remove predictors
* **Stepwise (sequential)**: combination of both

⚠️ These methods optimize prediction, **not causal interpretation**.

---

## Forward selection (default tuning)

```{r}
set.seed(123)
forwd.1 <- train(
  x = X_train, 
  y = y_train,
  method = "leapForward"
)
forwd.1
```

The model with the **lowest cross-validated RMSE** is chosen automatically.

```{r}
summary(forwd.1)

forwd.1$results$RMSE
which.min(forwd.1$results$RMSE)

coef(forwd.1$finalModel, which.min(forwd.1$results$RMSE))
```

---

### Student exercise 2: interpretation

```{r}
#| eval: false
# Why might forward selection fail
# to detect interacting predictors?
```

---

## Forward selection with expanded search

`caret()` defaults to searching over three sensible values. We can expand the search space by providing a grid.

```{r}
optimal.n <- expand.grid(nvmax = 1:10)

set.seed(123)
forwd.2 <-  train(
  x = X_train, 
  y = y_train,
  method = "leapForward",
  tuneGrid = optimal.n
)
forwd.2
```

The best model is still the one with 4 variables.


Extract selected predictors:

```{r}
nvchosen <- which.min(forwd.2$results$RMSE)

frd.names <- names(
  coef(forwd.2$finalModel, nvchosen)
)[-1]

forward <- list(
  vars = frd.names,
  rmse = min(forwd.2$results$RMSE)
)
```

---

## Backward elimination

```{r}
set.seed(123)
bckwrd <-  train(
  x = X_train, 
  y = y_train,
  method = "leapBackward",
  tuneGrid = optimal.n
)
bckwrd
```

```{r}
bckwrd.names <- names(
  coef(bckwrd$finalModel, which.min(bckwrd$results$RMSE))
)[-1]

backward <- list(
  vars = bckwrd.names,
  rmse = min(bckwrd$results$RMSE)
)
```

---

### Student exercise 3: comparison

```{r}
#| eval: false
# Why might forward and backward selection
# return the same model?
```

---

## Stepwise (sequential) selection

```{r}
set.seed(123)
seqsel <-  train(
  x = X_train, 
  y = y_train,
  method = "leapSeq",
  tuneGrid = optimal.n
)
seqsel
```

```{r}
seqsel.names <- names(
  coef(seqsel$finalModel, which.min(seqsel$results$RMSE))
)[-1]

stepwise <- list(
  vars = seqsel.names,
  rmse = min(seqsel$results$RMSE)
)
```

---

## Comparing selected models

```{r}
selection.list <- list(
  forward = forward,
  backward = backward,
  stepwise = stepwise
)

selection.list
```

### Interpretation

* Stepwise selection minimizes RMSE
* Forward/backward offer **greater parsimony**
* Small RMSE differences often favor simpler models

---

### Student exercise 4: model choice

```{r}
#| eval: false
# Which model would you deploy?
# Justify your answer using
# both RMSE and number of predictors.
```

---

## Validation-set performance

```{r}
Price.for <- predict(forwd.2, newdata = X_valid)
for_sel <- postResample(Price.for, y_valid)

Price.back <- predict(bckwrd, newdata = X_valid)
back_sel <- postResample(Price.back,  y_valid)

Price.seq <- predict(seqsel, newdata = X_valid)
seq_sel <- postResample(Price.seq,  y_valid)

rbind(for_sel, back_sel, seq_sel)
```

---

## Key takeaways

* Feature selection improves generalization
* Sequential methods are fast and interpretable
* RMSE differences must be weighed against complexity
* These methods precede **regularization**, not replace it

