---
title: "10-01-logistic"
format: html
editor: visual
error: true
echo: true
---

## Learning objectives

After completing this module, students should be able to:

-   Explain why linear regression is inappropriate for classification
-   Interpret logistic regression coefficients as log-odds and odds ratios
-   Fit logistic regression models using `train()`
-   Generate predicted probabilities from a logistic model
-   Understand how probability cutoffs relate to classification decisions

------------------------------------------------------------------------

Note: This is the updated project file.

## Ch 10: Logistic Regression

Classification refers to the idea of assigning a success or failure class (in the case of binary outcomes) to to an observation. This class assignment is based on generating a probability of the outcome for that case and then classifying the record using a probability cutoff value. Since, we need to generate probabilities, we can not use the traditional ordinary least squared regression approach there is nothing in the model that restricts the predictions to be between 0 and 1. Logistic regression is an approach that helps us get around that.

This file introduces logistical regression using the `caret()` syntax.

```{r}
#| echo: true
#| message: false
#| warning: false
library(tidyverse)
#library(gains)
library(caret)
library(yardstick)
library(broom)
library(mlbadata)
```

Let us first model the acceptance of personal loan as a function on income. Let `p` denote the probability of accepting a personal loan. The linear model to estimate the probability is given by:

$p = \beta_0 + \beta_1 income$

We will contrast this with a logistic function.

The *logistic response function* is given by:

$p = \frac{1}{1 + \exp^{-(\beta_0 + \beta_1 income)}}$

We will estimate the response from both the models and plot the resulting curve.

For estimation, we will continue the use of the `train()` function and the method will be `glm`.

```{r}

toy.df <- mlbadata::UniversalBank[,c(4,10)]
toy.df$Class <- factor(toy.df$Personal.Loan, 
                                levels = c(1,0),
                                labels = c("Yes", "No"))


# First, the linear regression
reg.1 <- lm(Personal.Loan ~ Income, data = toy.df)
tidy(reg.1)

```

Now logit

```{r}
Cntrl <- trainControl( method = "none",
                       classProbs = TRUE,
                      summaryFunction = twoClassSummary)

logit.1 <- train(Class ~ Income,
                 data = toy.df,
                 trControl = Cntrl,
                 method = "glm",
                 family = "binomial"
                 )

logit.1
```

In `caret()` the results are stored in the object `$finalModel`

```{r}
summary(logit.1$finalModel)

```

We can obtain this in the tidy format:

```{r}
tidy(logit.1$finalModel)
```

**Important**

Notice the coefficient on `Income`. It is negative in the above estimation but positive in the regression model. This is because `caret()` expects the first level to be the target level while the underlying algorithm that it runs, `glm()` assumes that the second level is the target level. So the results reported are actually modelling y = "No". When we use `augment()` or `predict()` to obtain probabilities after `train()`, we will still get the predictions for yes and no classes in the right manner. Keep this in mind however, when you interpret the estimation results.

Now, to obtain the predicted probabilities from the linear regression model and the logit model, we can use the `predict()` function.

```{r}
predict(reg.1, newdata = head(toy.df, 15))
```

We see that some of the probabilities are negative. This is not permissible.

In case of the logit model, we will see that the probabilities are constrained between 0 to 1.

```{r}
predict(logit.1, newdata = head(toy.df, 15), type = "prob")

```

To see the predicted class:

```{r}
predict(logit.1, newdata = head(toy.df, 15), type = "raw") |> as_tibble()
```

Plotting this will make this point clear. To obtain the predicted probably, we have have carry out some transformations. Below, we plot the predicted probabilities from the linear as well as the logit model

```{r}
# We first define the two functions

lin.mod <- function(x){
  reg.1$coefficients[1] + reg.1$coefficients[2]*x
}

#since the results are targeting "No" we subtract from 1 the prob to get
## prediction for class = 1
logit.mod <- function(x){
  (1- 1 / (1 + exp(-(logit.1$finalModel$coefficients[1] +  logit.1$finalModel$coefficients[2]*x))))
}

toy.df %>% 
  ggplot(aes(x = Income, y = Personal.Loan)) +
  geom_jitter(width=0.01, height=0.02, alpha=0.1) +
  geom_function(fun = lin.mod, color = "blue") +
  geom_function(fun = logit.mod, color = "red") +
  labs(x = "Income in '000s",
       y = "Personal Loan")
  
```

|  |
|----|
| \### Student Exercise 1 — Conceptual check Q: Why is the blue (linear) curve problematic as a probability model? |
| Q: List two concrete issues visible in the plot. |

What we notice is that while the linear probability model had probability estimates which were less than 0, the predicted probabilities in the logit model were between 0 to 1.

### Logistic Regression: full model

In general, the *logistic response function* is given by:

$$p = \frac{1}{1 + \exp^{-(\beta_0 + \beta_1 x_1 + \beta_2 x_2 \dots + \beta_K x_K)}}$$

If we define odds of success as: $$Odds( Y = 1) = \frac{p}{1-p}$$

Then, with some transformation, p equals: $$p = \frac{odds}{1+ odds}$$

Substituting the above expression into our logistic response function, we can obtain:

$$ Odds( Y = 1) = \exp^{\beta_0 + \beta_1 x_1 + \beta_2 x_2 \dots + \beta_K x_K} $$ Or, taking log on both sides, $$ \log(odds) = \beta_0 + \beta_1 x_1 + \beta_2 x_2 \dots + \beta_K x_K$$

This is the equation of the standard form of the logistic model that is estimated by `glm()` under the option of `family = "binomial"`.

Let us now estimate our model on the full dataset.

```{r}

bank.df <- mlbadata::UniversalBank
bank.df <- bank.df[ , -c(1, 5)]  # Drop ID and zip code columns.
# treat Education as categorical 
bank.df$Education <- factor(bank.df$Education, levels = c(1, 2, 3), 
                            labels = c("Undergrad", "Grad", "Adv/Prof"))

bank.df$Personal.Loan <- factor(bank.df$Personal.Loan, 
                                levels = c(1,0),
                                labels = c("Yes", "No"))

glimpse(bank.df)
# partition data
set.seed(1111)
train.index <- createDataPartition(bank.df$Personal.Loan, 
                                   p = 0.8,
                                   list = F)  
train.df <- bank.df[train.index, ]
test.df <- bank.df[-train.index, ]

```

Let's create full-rank dummy variables for the education category.

```{r}
dumvar <- dummyVars(Personal.Loan ~ .,
                    data = bank.df,
                    fullRank = TRUE)

```

Now, generate the training and testing dataset for X and y

```{r}
X_train <- predict(dumvar, newdata = train.df) |> as_tibble()
X_test <- predict(dumvar, newdata = test.df) |> as_tibble()

y_train <- train.df$Personal.Loan
y_test <- test.df$Personal.Loan
```

### Estimating the logsitical regression model

Here, we are not doing any tuning of hyper-parameters. So we will not use any resampling procedures. We will run our model and predict the class probabilities on the test set. We will then use some model evaluation metrics to assess the quality of the classification.

We will first define the `trainControl()` function.

```{r}
fitControl <- trainControl( method = "none",
                       classProbs = TRUE,
                      summaryFunction = twoClassSummary)                   
```

```{r}
logit.reg <- train(y = y_train,
                   x = X_train,
                   trControl = fitControl,
                   method = "glm",
                   family = "binomial"
                   )

summary(logit.reg)


```

The estimates give us the impact of a unit change in the x variable on *log odds* of **rejecting** a personal loan. For interpretation, we obtain the odds ratios.

```{r}
options(scipen = 999) #R default is (scipen = 0)
options(digits = 7) # R default is (digits = 7)
tidy(logit.reg$finalModel, exponentiate = T)


options(scipen = 0)
options(digits = 7)

#We can save the exponentiated coefficient estimates are 
result.1 <- tidy(logit.reg$finalModel, exponentiate = T)

result.1 |> 
  mutate(estimate = round(estimate,4),
         p.value = round(p.value, 4))
```

------------------------------------------------------------------------

### Student Exercise 2 — Interpretation

Q1. Pick any three continuous variables. Interpret the odds-ratios. Pay attention to the way `glm` expects the target class.

Q2. What is the interpretation of the estimates of the educational variables?

------------------------------------------------------------------------

A number greater than 1 indicates that the factor increases the odds of the sample being in the target class.

### Classifying the outcome for the train and test sets

We will assess the performance using both class-based and probability-based metrics.

Just as a recap, the following metrics can be obtained from `yardstick()`:

| **Metric**     | **Function**   | **Input Required**           |
|----------------|----------------|------------------------------|
| **Accuracy**   | `accuracy()`   | Predicted Class (`yes`/`no`) |
| **F1 Score**   | `f_meas()`     | Predicted Class (`yes`/`no`) |
| **Precision**  | `precision()`  | Predicted Class (`yes`/`no`) |
| **AUC-ROC**    | `roc_auc()`    | **Probability** (0.0 to 1.0) |
| **Lift Curve** | `lift_curve()` | **Probability** (0.0 to 1.0) |

Predictions: Remember the predictions are a probability. Give the option `type = "prob"` to obtain the probabilities. First the training set.

```{r}
predProb <- predict(logit.reg, type = "prob")
predClass <- predict(logit.reg,  type = "raw") 
train_result <- cbind(y_train, predProb, predClass)


```

1.  Confusion Matrix

```{r}
confusionMatrix(data = predClass, reference = y_train )
```

------------------------------------------------------------------------

### Student Exercise 3 — Interpretation

Q. What happens when we change the cutoff?

```{r}
confusionMatrix(data = factor(ifelse(predProb$Yes>0.25, "Yes", "No"),
                              levels = c("Yes", "No")),
                reference = y_train)
```

Now, increase the cut-off to 0.75

```{r}
#| eval: false
# Copy the code above and make the change in cutoff.
```

Q What do you observe?

------------------------------------------------------------------------

Now we work with the test set.

```{r}




predProb_test <- predict(logit.reg, newdata = X_test, type = "prob")
predClass_test <- predict(logit.reg, newdata = X_test, type = "raw")
result_test <- cbind(y_test, predProb_test, predClass_test)

confusionMatrix(data = predClass_test, reference = y_test)
```

Let's now obtain the lift curve and the ROC curve.

```{r}
lift_test <- result_test |> 
    lift_curve(truth = y_test, Yes)

autoplot(lift_test) +
  scale_x_continuous(breaks = seq(0,100,10))
```

```{r}
result_test |> 
  roc_auc(truth = y_test, Yes) 
```

To plot the ROC curve:

```{r}
roc_data <- result_test |> 
  roc_curve(truth = y_test, Yes) |> autoplot()
```
