---
title: "10-02 Penalized Logistic Regression with Cross-Validation"
format: html
editor: visual
error: true
echo: true
---

## Learning objectives

After completing this module, students should be able to:

-   Explain why regularization is useful in classification
-   Implement L1 (LASSO) Regularization, L2 (ridge) and Elastic Net
-   Use cross-validation to tune regularization strength, through grid and random
-   Compare logistic models using ROC, AUC, and lift curves

------------------------------------------------------------------------

## Assessing the performance of logistic regression models

In this module, we assess the performance of logistic regression models using the **caret framework**. We begin with a baseline (unpenalized) logistic regression and then extend it to a **penalized (LASSO) logistic regression** using cross-validation.

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(caret)
library(yardstick)
library(broom)
library(mlbadata)
```





Let us now estimate our model on the full dataset.

```{r}

bank.df <- mlbadata::UniversalBank
bank.df <- bank.df[ , -c(1, 5)]  # Drop ID and zip code columns.
# treat Education as categorical 
bank.df$Education <- factor(bank.df$Education, levels = c(1, 2, 3), 
                            labels = c("Undergrad", "Grad", "Adv/Prof"))

bank.df$Personal.Loan <- factor(bank.df$Personal.Loan, 
                                levels = c(1,0),
                                labels = c("Yes", "No"))

glimpse(bank.df)
# partition data
set.seed(1111)
train.index <- createDataPartition(bank.df$Personal.Loan, 
                                   p = 0.8,
                                   list = F)  
train.df <- bank.df[train.index, ]
test.df <- bank.df[-train.index, ]

```

Let's create full-rank dummy variables for the education category.

```{r}
dumvar <- dummyVars(Personal.Loan ~ .,
                    data = bank.df,
                    fullRank = TRUE)

```

Now, generate the training and testing dataset for X and y

```{r}
X_train <- predict(dumvar, newdata = train.df) |> as.data.frame()
X_test <- predict(dumvar, newdata = test.df) |> as.data.frame()

y_train <- train.df$Personal.Loan
y_test <- test.df$Personal.Loan
```

### Estimating the logsitical regression model

Here, we are not doing any tuning of hyper-parameters. So we will not use any resampling procedures. We will run our model and predict the class probabilities on the test set. We will then use some model evaluation metrics to assess the quality of the classification.

We will first define the `trainControl()` function.

```{r}
fitControl <- trainControl( method = "none",
                       classProbs = TRUE,
                      summaryFunction = twoClassSummary)                   
```

```{r}
logit.reg <- train(y = y_train,
                   x = X_train,
                   trControl = fitControl,
                   method = "glm",
                   family = "binomial"
                   )



```

------------------------------------------------------------------------




## Test-set predictions (baseline model)

```{r}
logit.reg_results <- bind_cols(
  predClass = predict(logit.reg, newdata = X_test, type = "raw"),
  predProb  = predict(logit.reg, newdata = X_test, type = "prob")$Yes,
  Actual    = y_test,
  model     = "logit.reg"
)
```

------------------------------------------------------------------------

## Penalized logistic regression (LASSO)

The baseline model includes all predictors. To reduce overfitting and encourage parsimony, we now apply **LASSO regularization**, which shrinks some coefficients exactly to zero.

We tune the penalty parameter ( \lambda ) using **10-fold cross-validation**.

```{r}
set.seed(1111)

fitCntrl <- trainControl(
  method = "cv",
  number = 10,
  allowParallel = TRUE,
  classProbs = TRUE,
  summaryFunction = twoClassSummary
)

lasso.search <- expand.grid(
  lambda = 10^(seq(-4, 2, length.out = 100)),
  alpha  = 1
)
```

```{r}
logit.l1 <- train(
  y = y_train,
  x = X_train,
  method = "glmnet",
  family = "binomial",
  trControl = fitCntrl,
  tuneGrid = lasso.search
)

logit.l1
```

------------------------------------------------------------------------

### Extracting coefficients from the tuned model

```{r}
coef(
  logit.l1$finalModel,
  logit.l1$bestTune$lambda
)
```

Odds ratios:

```{r}
exp(
  coef(
    logit.l1$finalModel,
    logit.l1$bestTune$lambda
  )
)
```

**Note:** Variables removed by LASSO have odds ratios equal to 1, indicating no effect.

------------------------------------------------------------------------

### Student Exercise 2 — Regularization insight

```{r}
#| eval: false

# Which variables were eliminated by the LASSO penalty?
# Why might this be desirable in a business setting?
```

------------------------------------------------------------------------

## Test-set predictions (penalized model)

```{r}
logit.l1_results <- bind_cols(
  predClass = predict(logit.l1, newdata = X_test, type = "raw"),
  predProb  = predict(logit.l1, newdata = X_test, type = "prob")$Yes,
  Actual    = y_test,
  model     = "logit.l1"
)

```

------------------------------------------------------------------------

### Elastic Net

Here, we will search for `$\alpha$` between 0 to 1. Recall 0: Ridge and 1: Lasso

We tune the penalty parameter ( \lambda ) using **10-fold cross-validation**.

```{r}
set.seed(1111)

fitCntrl <- trainControl(
  method = "cv",
  number = 10,
  allowParallel = TRUE,
  classProbs = TRUE,
  summaryFunction = twoClassSummary
)

enet.search <- expand.grid(
  lambda = 10^(seq(-4, 2, length.out = 100)),
  alpha  = seq(0,1,by = 0.1)
)
```

```{r}
logit.enet <- train(
  y = y_train,
  x = X_train,
  method = "glmnet",
  family = "binomial",
  trControl = fitCntrl,
  tuneGrid = enet.search
)

coef(logit.enet$finalModel, , s = logit.enet$bestTune$lambda)
```



## Test-set predictions (enet)

```{r}
logit.enet_results <- bind_cols(
  predClass = predict(logit.enet, newdata = X_test, type = "raw"),
  predProb  = predict(logit.enet, newdata = X_test, type = "prob")$Yes,
  Actual    = y_test,
  model     = "logit.enet"
)


```


### Random search.

Let us also do a random search for the hyperparameters. This is done by setting `search = "random"` in the `trainControl()` function and then setting the `tuneLength = n` where n is the number of values to search over.



```{r}
set.seed(1111)

fitCntrl <- trainControl(
  method = "cv",
  number = 10,
  search = "random",
  allowParallel = TRUE,
  classProbs = TRUE,
  summaryFunction = twoClassSummary
)


```

Now the random search model:

```{r}
logit.enetrand <- train(
  y = y_train,
  x = X_train,
  method = "glmnet",
  family = "binomial",
  trControl = fitCntrl,
  tuneLength = 100
)

summary(logit.enetrand)
```
To obtain the coefficient estimates from the best tuned model:

```{r}
coef(logit.enetrand$finalModel, s = logit.enetrand$bestTune$lambda)
```


```{r}
logit.enetrand_results <- bind_cols(
  predClass = predict(logit.enetrand, newdata = X_test, type = "raw"),
  predProb  = predict(logit.enetrand, newdata = X_test, type = "prob")$Yes,
  Actual    = y_test,
  model     = "logit.enetrand"
)

perf_results <- bind_rows(
  logit.reg_results,
  logit.l1_results,
  logit.enet_results,
  logit.enetrand_results)

```

## Model comparison: ROC, AUC, and lift

### Area under the curve (AUC)

```{r}
perf_results %>% 
  group_by(model) %>% 
  roc_auc(truth = Actual, predProb) |> arrange(-.estimate)
```

### ROC curves

```{r}
perf_results %>% 
  group_by(model) %>% 
  roc_curve(Actual, predProb) %>% 
  autoplot()
```

### Lift curves

```{r}
perf_results %>% 
  group_by(model) %>% 
  lift_curve(Actual, predProb) %>% 
  autoplot()
```

------------------------------------------------------------------------

### Student Exercise 3 — Model evaluation

```{r}
#| eval: false

# Compare the baseline and penalized models.
# Which performs better on AUC?
# Does the penalized model sacrifice much predictive power?
```

------------------------------------------------------------------------

## Wrap-up

-   Penalized logistic regression reduces overfitting
-   Cross-validation selects the appropriate penalty strength
-   LASSO performs implicit feature selection
-   Performance tradeoffs can be evaluated using ROC and lift curves

This framework generalizes directly to most other classification models to be discussed later.

## Final Step

Once we have identified our model, we will run the model on our entire dataset using the hyperparameter values from the best tuned result. The results from this model will be deployed to new data. We see that our enet model was the best model among the four algorithms that we ran. 

```{r}
logit.enet$bestTune
```

We will run this model on the entire dataset.

```{r}

X_full <- predict(dumvar, newdata = bank.df) |> as.data.frame()
y_full <- bank.df$Personal.Loan 

  
Cntrl <- trainControl(
  method = "none",
  allowParallel = TRUE
)

enetGrid <- expand.grid(
  alpha = logit.enet$bestTune$alpha,
  lambda = logit.enet$bestTune$lambda
)

final.model <- train(y = y_full,
                     x = X_full,
                    trControl = Cntrl,
                    tuneGrid = enetGrid,
                    method = "glmnet",
                    family = "binomial"
                     )

coef(final.model$finalModel, s = final.model$bestTune$lambda)
```

This `final.model` can be used to score new data.
